---
title: "Perturbing the Beltrami equations: a first-order analysis"
author: "Robin Hankin"
output: pdf_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```



\newcommand{\di}[2]{\frac{\partial{#1}}{\partial{#2}}}
\newcommand{\cx}{\cosh{\left(\frac{x-x_0}{a}\right)}}
\newcommand{\sx}{\sinh{\left(\frac{x-x_0}{a}\right)}}
\newcommand{\tx}{\tanh{\left(\frac{x-x_0}{a}\right)}}
\newcommand{\sech}{\operatorname{sech}}


Start with the Euler-Lagrange equations:

\begin{equation}
\frac{d}{dx}\left[\di{F}{y'}\right]=\di{F}{y}
\end{equation}

If $F$ is independent of $x$ we have the Beltrami equation:

\begin{equation}
F-y'\di{F}{y'}=C
\end{equation}


and consider a perturbation:

\begin{eqnarray}
y  &\longrightarrow&  y(x) + \epsilon z(x)\\
y' &\longrightarrow&  y'(x) + \epsilon z'(x)\\
F  &\longrightarrow&  F + \epsilon G(x,y,y')
\end{eqnarray}


\begin{eqnarray}
F-y'\di{F}{y'}=C\longrightarrow\\
F(x,y+\epsilon z,y'+\epsilon z')+\epsilon G(x,y+\epsilon z,y'+\epsilon z'-
(y'+\epsilon z')\di{}{y'}\left[F(x,y+\epsilon z,y'+\epsilon z')
+\epsilon G(x,y+\epsilon z,y'+\epsilon z')\right]\\
F(x,y+\epsilon z,y'+\epsilon z')+\epsilon G(x,y,y') -
(y'+\epsilon z')\di{}{y'}\left[F(x,y+\epsilon z,y'+\epsilon z')
+\epsilon G(x,y,y')\right]\\
F + \epsilon z\di{F}{y} + \epsilon z'\di{F}{y'}
+\epsilon G-
(y'+\epsilon z')\di{}{y'}\left[
F + \epsilon z\di{F}{y} + \epsilon z'\di{F}{y'}
+\epsilon G
\right]\\
F + \epsilon z\di{F}{y} + \epsilon z'\di{F}{y'}
+\epsilon G-
y'\di{}{y'}\left[
F + \epsilon z\di{F}{y} + \epsilon z'\di{F}{y'}
+\epsilon G
\right] -
\epsilon z'\di{F}{y'}\\
\epsilon z\di{F}{y} + \epsilon z'\di{F}{y'}
+\epsilon G -
y'\di{}{y'}\left[
\epsilon z\di{F}{y} + \epsilon z'\di{F}{y'}
+\epsilon G
\right] -
\epsilon z'\di{F}{y'}\\
z\di{F}{y} + z'\di{F}{y'}
+ G -
y'\di{}{y'}\left[
z\di{F}{y} + z'\di{F}{y'}
+  G
\right] -
z'\di{F}{y'}
\end{eqnarray}


(above, each line is equal to a constant).

Writing $H= z\di{F}{y} + z'\di{F}{y'}+ G$ we have

\begin{equation}
H-y'\di{H}{y'}-z'\di{F}{y'}=C\end{equation}

might make the catenary easier, but only if $\phi(x,y)$ is independent
of $x$.  Here we have $F=(y+\lambda)\sqrt{1+y'^2}$ and
$G=\phi(y)\sqrt{1+y'^2}$ so

\begin{eqnarray}
H &=& z\di{F}{y} + z'\di{F}{y'}+ G\\
&=& z\sqrt{1+y'^2} + \frac{z'y'(y + \lambda)}{\sqrt{1+y'^2}} + \phi(y)\sqrt{1+y'^2}
\end{eqnarray}

So we have

\begin{eqnarray}
\di{H}{y'} &=& \frac{zy'}{\sqrt{1+y'^2}} + \frac{z'(y+\lambda)}{(1+y'^2)^{3/2}} + \frac{y'\phi(y)}{\sqrt{1+y'^2}}\\
&=& \frac{y'(z+\phi(y))}{\sqrt{1+y'^2}} + \frac{z'(y+\lambda)}{(1+y'^2)^{3/2}}
\end{eqnarray}

\begin{eqnarray}
H-y'\di{H}{y'}-z'\di{F}{y'}\\
\underbrace{z\sqrt{1+y'^2} + \frac{z'y'(y + \lambda)}{\sqrt{1+y'^2}}+ \phi\sqrt{1+y'^2}}_{H}
- \frac{y'^2(z+\phi)}{\sqrt{1+y'^2}} - \frac{y'z'(y+\lambda)}{(1+y'^2)^{3/2}}
- \frac{z'y'(y+\lambda)}{\sqrt{1+y'^2}}\\
z\sqrt{1+y'^2} + \phi\sqrt{1+y'^2}
- \frac{y'^2(z+\phi)}{\sqrt{1+y'^2}} - \frac{y'z'(y+\lambda)}{(1+y'^2)^{3/2}}\\
(z+\phi)\sqrt{1+y'^2} 
- \frac{y'^2(z+\phi)}{\sqrt{1+y'^2}} - \frac{y'z'(y+\lambda)}{(1+y'^2)^{3/2}}\\
(z+\phi)\left(\sqrt{1+y'^2}  - \frac{y'^2}{\sqrt{1+y'^2}}\right) - \frac{y'z'(y+\lambda)}{(1+y'^2)^{3/2}}\\
\frac{z+\phi}{\sqrt{1+y'^2}} - \frac{y'z'(y+\lambda)}{(1+y'^2)^{3/2}}
\end{eqnarray}

The above are all equal to a constant $C$. So


\begin{eqnarray}
z+\phi - \frac{y'z'(y+\lambda)}{1+y'^2} = C\sqrt{1+y'^2}
\end{eqnarray}

and at this point we can substitute for $y=\cosh x$ and $y'=\sinh x$:

\begin{eqnarray}
z+\phi - \frac{z'\sinh x(\cosh x+\lambda)}{\cosh^2x} = C\cosh x\\
z+\phi - C\cosh x =
\frac{z'\sinh x(\cosh x+\lambda)}{\cosh^2x}\\
z'= \frac{(z+\phi - C\cosh x)\cosh^2x}{\sinh x(\cosh x+\lambda)}
\end{eqnarray}


The problem is that if $x=0$, $\sinh x=0$ so $z'$ becomes infinite???

## Numerical work


```{r useode}
library("deSolve") 
Lorenz <- function(t, state, parameters) {
    with(as.list(c(state, parameters)), {

        phi <- cosh(t)^2  # y^2

        numerator <- (Z + phi - C*cosh(t)) * cosh(t)^2
        denominator <- sinh(t) * (cosh(t) + lambda)

        dZ  <- numerator/denominator
        dextra_length <- -Z/cosh(t)^2
        list(c(dZ, dextra_length))
    })
}

parameters <- c(lambda = 3, C=0.5)
state      <- c(Z = 0, extra_length=0)  # initial conditions
times      <- seq(-2, 2, by = 0.01)

out <- ode(y = state, times = times, func = Lorenz, parms = parameters)
```

```{r showode}
head(out)
tail(out)
x <- out[,1]
y <- out[,2]
```


```{r badnessdef}
badness <- function(unk, give=FALSE, byval=1){
   unk <- unname(unk)
   parameters <- c(lambda = unk[1], C=unk[2])
   state <- c(Z = 0, extra_length=0)  # initial conditions
   times      <- seq(-2, 2, by = byval)

   out <- ode(y = state, times = times, func = Lorenz, parms = parameters)
   Z_bad <- out[5,2]
   L_bad <- out[5,3]
   if(give){return(out)}
   return(Z_bad^2 + L_bad^2)
}

badness(c(1,2), give=TRUE)
badness(c(1,2), give=FALSE)
```

```{r, findbad, cache=TRUE}
  lambda <- seq(from=0, to=7, by=0.1)
  C      <- seq(from=-1, to=1, by=0.1)

jj <- as.matrix(expand.grid(lambda,C))
bad <- apply(jj,1,badness)
```

```{r}
bad <- matrix(bad, length(lambda), length(C))
contour(lambda, C, bad)
filled.contour(lambda, C, bad)
matplot(bad)
```


```{r, optimbad, cache=TRUE}
optim(c(5,0),badness)
```




```{r}
out <- badness(c(5.87,-0.73),give=TRUE,byval=0.01)
head(out)
plot(out[,1],out[,2],type='b')
```
