---
title: "Perturbing the catenary: a first-order analysis"
author: "Robin Hankin"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The catenary.

\newcommand{\di}[2]{\frac{\partial{#1}}{\partial{#2}}}
\newcommand{\cx}{\cosh{\left(\frac{x-x_0}{a}\right)}}
\newcommand{\sx}{\sinh{\left(\frac{x-x_0}{a}\right)}}
\newcommand{\tx}{\tanh{\left(\frac{x-x_0}{a}\right)}}
\newcommand{\sech}{\operatorname{sech}}

We start with functional $F = \sqrt{1+y'^2}(y+\lambda)$ and
Euler-Lagrange equation

\begin{equation}
\frac{d}{dx}\di{F}{y'} -\di{F}{y}=0
\end{equation}

We have $\di{F}{y'} = \frac{y'(y+\lambda)}{\sqrt{1+y'^2}}$ and 
$\di{F}{y} = \sqrt{1+y'^2}$ so

\begin{eqnarray}
\frac{d}{dx}\left[\frac{y'(y+\lambda)}{\sqrt{1+y'^2}}\right] &=& \sqrt{1+y'^2}\nonumber\\
\frac{y''(y + \lambda)}{(1+y'^2)^{3/2}} + \frac{y'^2}{\sqrt{1+y'^2}}        &=& \sqrt{1+y'^2}
\end{eqnarray}

This gives $y = \operatorname{cosh} x$.



# Perturbing the catenary.

We now consider $Y=Y(x) = y(x) + \epsilon z(x)$, a slightly perturbed
solution to the EL equations above.  The first-order corrective term
gives functional

\begin{equation}
\overline{F} = \sqrt{1+Y'^2}(Y + \Lambda + \epsilon\phi(x,Y))
\end{equation}

where $\epsilon$ is small.  E-L becomes:

\begin{equation}
\frac{d}{dx}\di{\overline{F}}{Y'} -\di{\overline{F}}{Y}=0
\end{equation}

And we have

\begin{equation}
\di{\overline{F}}{Y'}
= \frac{Y'}{\sqrt{1+Y'^2}}(Y + \Lambda + \epsilon\phi(x,y))\qquad
\di{\overline{F}}{Y} = {\sqrt{1+Y'^2}}(1 + \epsilon_y\phi(x,y))
\end{equation}

Thus (working to first order) [also recalling that $\phi=\phi(x,y)$]

\begin{eqnarray}
\frac{d}{dx}\left[\di{\overline{F}}{Y'}\right]
&=& \frac{d}{dx}\left[\frac{Y'}{\sqrt{1+Y'^2}}(Y + \Lambda + \epsilon\phi)\right]\\
&=& \frac{Y''}{(1+Y'^2)^{3/2}}(Y + \Lambda + \epsilon\phi) + 
        \frac{Y'}{\sqrt{1+Y'^2}}(y'+\epsilon z' + \epsilon(\phi_x + y'\phi_y))\\
&=& \frac{y'' + \epsilon z''}{(1+(y'+\epsilon z')^2)^{3/2}}(y + \epsilon z + \Lambda + \epsilon\phi) + 
        \frac{y' + \epsilon z'}{\sqrt{1+(y'+\epsilon z')^2}}(y'+\epsilon z' + \epsilon(\phi_x + y'\phi_y))\\
&=& \left(\frac{y''}{(1+(y'+\epsilon z')^2)^{3/2}} + 
\frac{\epsilon z''}{(1+(y'+\epsilon z')^2)^{3/2}}\right)(y + \epsilon z + \Lambda + \epsilon\phi)\nonumber\\
&{}& \qquad + \frac{y' + \epsilon z'}{\sqrt{1+(y'+\epsilon z')^2}}(y'+\epsilon z' + \epsilon(\phi_x + y'\phi_y))\\
&=& \left(\frac{y''}{(1+(y'+\epsilon z')^2)^{3/2}} + 
\frac{\epsilon z''}{(1+(y'+\epsilon z')^2)^{3/2}}\right)(y(x) + \epsilon z + \Lambda + \epsilon\phi)\nonumber\\
&{}& \qquad + \left(\frac{y'}{\sqrt{1+y'^2}} +
\epsilon\frac{z'}{(1+y'^2)^{3/2}} + \epsilon\frac{z'}{\sqrt{1+y'^2}}\right)(y'+\epsilon z' + \epsilon(\phi_x + y'\phi_y))\\
&=& \left(\frac{y''}{(1+(y'+\epsilon z')^2)^{3/2}} + 
\frac{\epsilon z''}{(1+(y'+\epsilon z')^2)^{3/2}}\right)(y +\Lambda + \epsilon(z + \phi))\nonumber\\
&{}& \qquad + \left(\frac{y'}{\sqrt{1+y'^2}} +
\epsilon\frac{z'}{(1+y'^2)^{3/2}} +
\epsilon\frac{z'}{\sqrt{1+y'^2}}\right)(y'+\epsilon (z' + \phi_x + y'\phi_y))\\
&=& \left(\frac{y''}{(1+(y'+\epsilon z')^2)^{3/2}} + 
\frac{\epsilon z''}{(1+(y'+\epsilon z')^2)^{3/2}}\right)(y +\Lambda + \epsilon(z + \phi))\nonumber\\
&{}& \qquad + \left(\frac{y'}{\sqrt{1+y'^2}} +
\epsilon\left(\frac{z'}{(1+y'^2)^{3/2}} + \frac{z'}{\sqrt{1+y'^2}}\right)\right)(y'+\epsilon (z' + \phi_x + y'\phi_y))\\
&=& \left(\frac{y''}{(1+(y'+\epsilon z')^2)^{3/2}} + 
\frac{\epsilon z''}{(1+(y'+\epsilon z')^2)^{3/2}}\right)(y +\Lambda + \epsilon(z + \phi))\nonumber\\
&{}& \qquad + \frac{y'^2}{\sqrt{1+y'^2}} +
\epsilon\left(\frac{y'z'}{(1+y'^2)^{3/2}} + \frac{y'z'}{\sqrt{1+y'^2}}
 + \frac{y'(z' + \phi_x + y'\phi_y)}{\sqrt{1+y'^2}}
\right)\\
&=& \left(\frac{y''}{(1+(y'+\epsilon z')^2)^{3/2}} + 
\frac{\epsilon z''}{(1+y'^2)^{3/2}}\right)(y +\Lambda + \epsilon(z + \phi))\nonumber\\
&{}& \qquad + \frac{y'^2}{\sqrt{1+y'^2}} +
\epsilon\left(\frac{y'z'}{(1+y'^2)^{3/2}} + \frac{y'z'}{\sqrt{1+y'^2}}
 + \frac{y'(z' + \phi_x + y'\phi_y)}{\sqrt{1+y'^2}}
\right)\\
&=& \left(
\frac{y''}{(1+y')^2)^{3/2}} -3\epsilon\frac{y''z'}{(1+y'^2)^{5/2}} + 
\frac{\epsilon z''}{(1+y'^2)^{3/2}}\right)(y +\Lambda + \epsilon(z + \phi))\nonumber\\
&{}& \qquad + \frac{y'^2}{\sqrt{1+y'^2}} +
\epsilon\left(\frac{y'z'}{(1+y'^2)^{3/2}} + \frac{y'z'}{\sqrt{1+y'^2}}
 + \frac{y'(z' + \phi_x + y'\phi_y)}{\sqrt{1+y'^2}}
\right)\\
&=& \left(
\frac{y''}{(1+y')^2)^{3/2}} +\epsilon\left(\frac{-3y''z'}{(1+y'^2)^{5/2}} + 
\frac{z''}{(1+y'^2)^{3/2}}\right)
\right)(y +\Lambda + \epsilon(z + \phi))\nonumber\\
&{}& \qquad + \frac{y'^2}{\sqrt{1+y'^2}} +
\epsilon\left(\frac{y'z'}{(1+y'^2)^{3/2}} + \frac{y'z'}{\sqrt{1+y'^2}}
 + \frac{y'(z' + \phi_x + y'\phi_y)}{\sqrt{1+y'^2}}
\right)\\
&=&  \frac{y''(y+\Lambda)}{(1+y')^2)^{3/2}} +
    \epsilon\left(
	    \frac{-3y''(y+\Lambda)z'}{(1+y'^2)^{5/2}} + 
        \frac{z''(y+\Lambda)}{(1+y'^2)^{3/2}} +
		\frac{y''(z+\phi)}{(1+y'^2)^{3/2}}
        \right)\\
&{}& \qquad + \frac{y'^2}{\sqrt{1+y'^2}} +
\epsilon\left(\frac{y'z'}{(1+y'^2)^{3/2}} + \frac{y'z'}{\sqrt{1+y'^2}}
 + \frac{y'(z' + \phi_x + y'\phi_y)}{\sqrt{1+y'^2}}
\right)\\
&=&  \frac{y''(y+\Lambda)}{(1+y')^2)^{3/2}} +
    \epsilon\left(
	    \frac{-3y''(y+\Lambda)z'}{(1+y'^2)^{5/2}} + 
        \frac{z''(y+\Lambda)}{(1+y'^2)^{3/2}} +
		\frac{y''(z+\phi)}{(1+y'^2)^{3/2}}
        \right)\\
&{}& \qquad + \frac{y'^2}{\sqrt{1+y'^2}} +
\epsilon\left(\frac{y'z'}{(1+y'^2)^{3/2}} + \frac{y'z'}{\sqrt{1+y'^2}}
 + \frac{y'(z' + \phi_x + y'\phi_y)}{\sqrt{1+y'^2}}
\right)
\end{eqnarray}

where $\phi=\phi(x,y)$, $z=z(x)$ and $z'=z'(x)$.  Now the right hand side:

\begin{eqnarray}
RHS &=& \di{\overline{F}}{Y}\\
&=& \frac{\partial}{\partial Y}\left[\sqrt{1+Y'^2}(Y + \Lambda + \epsilon\phi(x,Y))\right]\\
&=& \sqrt{1+Y'^2}\left(1 + \epsilon\phi_Y(x,Y))\right)\\
&=& \sqrt{1+Y'^2} + \epsilon\phi_y(x,y)\sqrt{1+Y'^2}
\end{eqnarray}

But we know that $\sqrt{1+Y'^2} = \sqrt{1+(y'+\epsilon z')^2} =
\sqrt{1+y'^2} + \epsilon\frac{y'z'}{\sqrt{1+y'^2}}$, to first order.  So
we have, again to first order

\begin{eqnarray}
RHS
&=& \sqrt{1+y'^2} + \epsilon\frac{y'z'}{\sqrt{1+y'^2}}  + \epsilon\phi_y(x,y)\sqrt{1+Y'^2}\\
&=& \sqrt{1+y'^2} + \epsilon\left(\frac{y'z'}{\sqrt{1+y'^2}}  + \phi_y(x,y)\sqrt{1+Y'^2}\right)\\
&=& \sqrt{1+y'^2} + \epsilon\left(\frac{y'z'}{\sqrt{1+y'^2}}  + \phi_y(x,y)\sqrt{1+y'^2}\right)
\end{eqnarray}
 

So

\begin{eqnarray}
&{}& \frac{y''(y+\Lambda)}{(1+y'^2)^{3/2}} + \frac{y'}{\sqrt{1+y'^2}}\nonumber\\
&+& \epsilon\left(
    \frac{y''(z+\phi)}{(1+y'^2)^{3/2}}
-  3\frac{(y+\Lambda)y''z'}{(1+y'^2)^{5/2}}
+   \frac{z''(y+\Lambda)}{(1+y'^2)^{3/2}}
+   \frac{y'z'}{(1+y'^2)^{3/2}}
+   \frac{y'z'}{\sqrt{1+y'^2}}
+   \frac{y'(\phi_x + y'\phi_y)}{\sqrt{1+y'^2}}
\right) \nonumber\\
&=& \sqrt{1+y'^2} + \epsilon\left(\frac{y'z'}{\sqrt{1+y'^2}}  + \phi_y(x,y)\sqrt{1+y'^2}\right)\label{lhsrhs}
\end{eqnarray}

In equation \ref{lhsrhs} the zeroth order term (that is, $\epsilon^0$)
is just the Euler-Lagrange equation for the unperturbed functional.
The first-order term is (after cancelling one of the terms):

\begin{eqnarray}
    \frac{y''(z+\phi)}{(1+y'^2)^{3/2}}
-  3\frac{(y+\Lambda)y''z'}{(1+y'^2)^{5/2}}
+   \frac{z''(y+\Lambda)}{(1+y'^2)^{3/2}}
+   \frac{y'z'}{(1+y'^2)^{3/2}}
+   \frac{y'(\phi_x + y'\phi_y)}{\sqrt{1+y'^2}}
=
\phi_y\sqrt{1+y'^2}
\end{eqnarray}

\begin{eqnarray}
    \frac{y''(z+\phi) + z''(y+\Lambda) + y'z'}{(1+y'^2)^{3/2}}
-  3\frac{(y+\Lambda)y''z'}{(1+y'^2)^{5/2}}
+   \frac{y'(\phi_x + y'\phi_y)}{\sqrt{1+y'^2}}
=
\phi_y\sqrt{1+y'^2}\label{firstorderunsub}
\end{eqnarray}

Using a standard catenary, that is, $y=\cosh x$, $y'=\sinh x$,
$y''=\cosh x$ [noting that $\sqrt{1+y'^2}=\cosh x$], we have

\begin{eqnarray}
  \frac{\cosh x(z + \phi) + z''(\cosh x + \Lambda) + z'\sinh x}{\cosh^3x}
-3\frac{z(\cosh x + \Lambda)\cosh x}{\cosh^5 x}
+ \frac{\sinh x(\phi_x + \phi_y\sinh x)}{\cosh x}
= \phi_y\cosh x
\end{eqnarray}

Now substitute $\Lambda=0$:

\begin{eqnarray}
  \frac{\cosh x(z + \phi) + z''\cosh x + z'\sinh x}{\cosh^3x}
-\frac{3z}{\cosh^3 x}
+ \frac{\sinh x(\phi_x + \phi_y\sinh x)}{\cosh x}
= \phi_y\cosh x
\end{eqnarray}

\begin{eqnarray}
  \frac{\cosh x(z + \phi) + z''\cosh x + z'\sinh x - 3z}{\cosh^3x}
+ \frac{\sinh x(\phi_x + \phi_y\sinh x)}{\cosh x}
= \phi_y\cosh x
\end{eqnarray}

\begin{eqnarray}
  \frac{z''\cosh x + z'\sinh x  + z(\cosh x - 3) + \phi\cosh x}{\cosh^3x}
+ \frac{\sinh x(\phi_x + \phi_y\sinh x)}{\cosh x}
= \phi_y\cosh x
\end{eqnarray}

\begin{eqnarray}
  \frac{z''\cosh x + z'\sinh x  + z(\cosh x - 3) + \phi\cosh x}{\cosh^3x}
= \phi_y\cosh x - \frac{\sinh x(\phi_x + \phi_y\sinh x)}{\cosh x}
\end{eqnarray}

\begin{eqnarray}
z''\cosh x + z'\sinh x  + z(\cosh x - 3) + \phi\cosh x
= \phi_y\cosh^4 x - \cosh^2 x\sinh x(\phi_x + \phi_y\sinh x)
\end{eqnarray}

\begin{eqnarray}
z''\cosh x = 
  \phi_y\cosh^4 x 
- \cosh^2 x\sinh x(\phi_x + \phi_y\sinh x) 
- (z'\sinh x  + z(\cosh x - 3) + \phi\cosh x)
\end{eqnarray}

\begin{eqnarray}
z''\cosh x = 
-z'\sinh x 
+z(3 - \cosh x)
+ (
\phi_y\cosh^4 x - \phi\cosh x - 
\cosh^2 x\sinh x(\phi_x + \phi_y\sinh x)
)
\end{eqnarray}

\begin{eqnarray}
z'' = 
-z'\tanh x 
+ z(3\sech x - 1)
+ \phi_y\cosh^3 x - \phi - \cosh x\sinh x(\phi_x + \phi_y\sinh x)
\end{eqnarray}



# Numerical  work


```{r odesolver}
library("deSolve") 
sech <- function(x){1/cosh(x)}
Lorenz <- function(t, state, parameters) {
    with(as.list(c(state, parameters)), {
        dZ <- Zdash

        phi <- t
        phi_y  <- 0
        phi_x <- 1

        t1 <- -Zdash * tanh(t)
        t2 <- Z * (3/cosh(t) - 1)
        t3 <- phi_y * cosh(t)^3 - phi + lambda - cosh(t) * sinh(t) * (phi_x + phi_y*sinh(t))

        dZdash <- t1 + t2 + t3
        dextra_length <- Zdash * tanh(t)
        dextra_length2 <- -Z * sech(t)^2
        list(c(dZ, dZdash, dextra_length, dextra_length2))
    })
}

parameters <- c(lambda = 0)
state      <- c(Z = 0, Zdash = 0, extra_length = 0, extra_length2 = 0)  # initial conditions
times      <- seq(-2, 2, by = 0.05)

out <- ode(y = state, times = times, func = Lorenz, parms = parameters)
```

```{r showode}
head(out)
tail(out)
x <- out[, 1]
y <- out[, 2]
plot(x, y, type = 'l')
grid()
```


```{r defodefunc}
f <- function(unknown, give=FALSE){
    lambda        <- unknown[1]
    Zdash_initial <- unknown[2]
    times <- seq(from = -2, to = 2, by = 1)
    parameters <- c(lambda = lambda)
    state <-  # initial conditions
        c(Z = 0, Zdash = Zdash_initial, extra_length=0, extra_length2=0) 
    out <- ode(y = state, times = times, func = Lorenz, parms = parameters)
    final_disp <- out[5,2]
    extra_length <- out[5,4]
    badness <- final_disp^2 + extra_length^2
    if(give){
        return(out)
    } else {
        return(badness)
    } 
}
```

```{r}
f(c(2, 1.0))
f(c(2, 1.1))
```
   
   
```{r find2dopt, cache=TRUE}
jj <- optim(c(0,0),f)
jj
```

```{r usef, cache=TRUE}
v1 <- seq(from = -1, to =  1, by = 0.01)
v2 <- seq(from = -6, to = -4, by = 0.01)
M <- as.matrix(expand.grid(v1,v2))
head(M)
bad <- sapply(seq_len(nrow(M)),function(i){f(unname(M[i,]))})
```

```{r contbad}
M <- matrix(bad,length(v1),length(v2))
M <- pmin(M,10)
contour(v1, v2, M)
grid()
filled.contour(M)
```

next chunk

	
```{r}
jj$par
parameters <- c(lambda = jj$par[1])
state      <- c(Z = 0, Zdash = jj$par[2], extra_length=0, extra_length2=0)  # initial conditions
times      <- seq(-2, 2, by = 0.01)

out <- ode(y = state, times = times, func = Lorenz, parms = parameters)
head(out)
tail(out)
x  <- out[,1]
z <- out[,2]
y0 <- cosh(x)

plot(x, z, type="l")
grid()
plot(x, y0,type="l")
points(x, y0+z/10,type="l",col="red")
```


OK we have two points that are not the same but both satisfy the
boundary conditions and indeed the length constraint.


```{r}
p1 <- jj$par
p2 <- c(1,-5.5)
```

```{r}
show <- function(p){
    parameters <- c(lambda = p[1])
    state      <- c(Z = 0, Zdash = p[2], extra_length=0, extra_length2=0)  # initial conditions
    times      <- seq(-2, 2, by = 0.01)
    ode(y = state, times = times, func = Lorenz, parms = parameters)
}


jj1 <- show(p1)
jj2 <- show(p2)

head(jj1)

plot(times,jj1[,2])
plot(times,jj2[,2])
```

Now one-dimensional optimization, using `optimize()` to find the best
value for the initial slope.


```{r}
bestslope <- function(lambda){
    badness <- function(slope){f(unknown = c(lambda, slope))}
    optimize(badness, interval = c(-8,8))
    }

bestslope(-1)
bestslope(0)
bestslope(1)
```

```{r showbest}
thinkbest <- c(0,bestslope(0)$minimum)
jj <- show(thinkbest)
plot(jj[,2])
tail(jj)
```


```{r}
getobj <- function(lambda){bestslope(lambda)$objective}
L <- seq(from=-0.5,to=0.5,by=0.1)
OBJ <- sapply(L,getobj)
```

```{r}
plot(L,OBJ)
```



```{r}
parameters <- c(lambda = 0)
state      <- c(Z = 0, Zdash = -4.864535, extra_length=0, extra_length2=0)  # initial conditions
times      <- seq(-2, 2, by = 0.01)

out <- ode(y = state, times = times, func = Lorenz, parms = parameters)
head(out)
tail(out)
x  <- out[,1]
z <- out[,2]
y0 <- cosh(x)

plot(x, z,type="l")
plot(x, y0,type="l")
points(x, y0+z/5,type="l",col="red")
```
