---
-title: "Perturbing the Euler-Lagrange equations for calculus of variations: a first-order analysis"
author: "Robin Hankin"
output: pdf_document
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

\newcommand{\di}[2]{\frac{\partial{#1}}{\partial{#2}}}
\newcommand{\cx}{\cosh{\left(\frac{x-x_0}{a}\right)}}
\newcommand{\sx}{\sinh{\left(\frac{x-x_0}{a}\right)}}
\newcommand{\tx}{\tanh{\left(\frac{x-x_0}{a}\right)}}
\newcommand{\el}[1]{\frac{d}{dx}\di{#1}{y'} - \di{#1}{y}}

Consider a functional $F=F(x,y,y';\lambda)$ and a perturbation:

\begin{eqnarray}
y  &\longrightarrow&  y(x) + \epsilon z(x)\\
y' &\longrightarrow&  y'(x) + \epsilon z'(x)\\
F  &\longrightarrow&  F + \epsilon G(x,y,y')
\end{eqnarray}


\begin{eqnarray}
\frac{d}{dx}\di{F}{y'}=\di{F}{y}\nonumber\longrightarrow\\
\frac{d}{dx}\left[\di{}{y'}\left(F(x,y+\epsilon z,y'+\epsilon z')+\epsilon G(x,y+\epsilon z,y'+\epsilon z')\right)\right] =\\
\di{}{y}\left[F(x,y+\epsilon z,y'+\epsilon z')
+\epsilon G(x,y+\epsilon z,y'+\epsilon z')\right]
\end{eqnarray}

Above we write $y$ for $y(x)$ etc.  We have to remember that
$y,y',z,z'$ are functions of $x$.  Now neglect terms of order
$\epsilon^2$ and above, and continue to neglect them:

\begin{equation}
\left(\el{G}\right) + \left(z\di{}{y} + z'\di{}{y'} + z''\di{}{y''}\right)
\left(\el{F}\right)=0
\end{equation}


# Examples.

First, straight lines as in geometrical optics

\begin{equation}
F=\sqrt{1+y'^2},\qquad G = \phi(x,y)\sqrt{1+y'^2}
\end{equation}

Here we need

\begin{eqnarray}
\el{F} &=& \frac{d}{dt}\di{F}{y'}\nonumber\\
       &=& \frac{d}{dt}\frac{y'}{\sqrt{1+y'^2}}\nonumber\\
       &=& \frac{y''}{(1+y'^2)^{3/2}}
\end{eqnarray}

and 
\begin{eqnarray}
\el{G} &=& \frac{d}{dt}\di{G}{y'} - \phi_y\sqrt{1+y'^2}\nonumber\\
       &=& \frac{d}{dt}\frac{y'\phi}{\sqrt{1+y'^2}} - \phi_y\sqrt{1+y'^2}\nonumber\\
       &=& \frac{y''\phi}{(1+y'^2)^{3/2}} + \frac{y'(\phi_x + y'\phi_y)}{\sqrt{1+y'^2}} - \phi_y\sqrt{1+y'^2}\nonumber\\
       &=& \frac{y''\phi}{(1+y'^2)^{3/2}} + \frac{1}{\sqrt{1+y'^2}}\left(y'(\phi_x + y'\phi_y) - \phi_y-\phi_yy^2\right)\nonumber\\
       &=& \frac{y''\phi}{(1+y'^2)^{3/2}} + \frac{y'\phi_x  -\phi_y}{\sqrt{1+y'^2}}
\end{eqnarray}


\begin{eqnarray}
\left(\el{G}\right) + \left(z\di{}{y} + z'\di{}{y'} + z''\di{}{y''}\right)
\left(\el{F}\right) &=& 0\\
\frac{y''\phi}{(1+y'^2)^{3/2}} + \frac{y'\phi_x - \phi_y}{\sqrt{1+y'^2}} +
\left(z\di{}{y} + z'\di{}{y'} + z''\di{}{y''}\right)
\left(\frac{y''}{(1+y'^2)^{3/2}}\right)\\
\frac{y''\phi}{(1+y'^2)^{3/2}} + \frac{y'\phi_x - \phi_y}{\sqrt{1+y'^2}} +
\left(z'\di{}{y'} + z''\di{}{y''}\right)
\left(\frac{y''}{(1+y'^2)^{3/2}}\right)\\
\frac{y''\phi}{(1+y'^2)^{3/2}} + \frac{y'\phi_x - \phi_y}{\sqrt{1+y'^2}} +
\frac{-3y'y''z'}{(1+y'^2)^{5/2}} + \frac{z''}{(1+y'^2)^{3/2}} &=&0
\end{eqnarray}


At this point we observe that $y=a+sx$ [ "$s$" for slope] and so $y'=s$ and $y''=0$:

\begin{eqnarray}
\frac{y'\phi_x - \phi_y}{\sqrt{1+y'^2}} + \frac{z''}{(1+y'^2)^{3/2}} = 0\\
\frac{s\phi_x - \phi_y}{\sqrt{1+s^2}} + \frac{z''}{(1+s^2)^{3/2}} = 0
\end{eqnarray}

simplifying,

\begin{equation}
z'' = (1+s^2)(\phi_y -s\phi_x)
\end{equation}

## Numerical work: refractive index


```{r odedefiner}
library("deSolve") 
Lorenz <- function(t, state, parameters) {
    with(as.list(c(state, parameters)), {
        dZ <- Zdash

        x <- t
        y <- t*s
        
        jj <- exp(-((x-x0)^2 + (y-y0)^2)/r^2)
        phi <- jj
        phi_x <- -2*jj*(x-x0)/r^2
        phi_y <- -2*jj*(y-y0)/r^2

        dZdash <- (1+s^2)*(phi_y - s*phi_x)

        list(c(dZ, dZdash))
    })
}

```{r odesolverrr, cache = FALSE}
parameters <- c(s = 0.1,            # "s" for slope
                 x0=5, y0=0,  r=1)  # (x0,y0) = center; radius r
state      <- c(Z = 0, Zdash = 0)   # initial conditions
times      <- seq(0, 10, by = 0.1)
out <- ode(y = state, times = times, func = Lorenz, parms = parameters)
```

```{r showode, cache=FALSE}
head(out)
tail(out)
x <- out[,1]
y <- out[,2]
plot(x,y,type='l')
grid()
```

```{r definerayfunc}
ray <- function(s, epsilon = 0.2, give = FALSE){  # s for slope
    parameters <- c(s=s, x0=5, y0=0, r=1)
    state      <- c(Z = 0, Zdash = 0)   # initial conditions
    times      <- seq(0, 10, by = 0.1)
    out <- ode(y = state, times = times, func = Lorenz, parms = parameters)
    if(give){return(out)}
    x <- out[,1]
    y <- x*s + out[,2]*epsilon
    return(cbind(x,y))
}
```


```{r, plotdelta, cache=FALSE}
plot(NA, xlim=c(0, 10), ylim=c(0, 1), asp=1)
theta <- seq(from=0, to = 2*pi, len=100)
points(5 + cos(theta), sin(theta), type="l",col="gray")
for(i in seq(from=0, to=0.4, len=15)){
    points(ray(i), type='l')
}

```

```{r, plotdeltacloseup, cache=FALSE}
plot(NA, xlim=c(6,8), ylim=c(-0.2, 0.2), asp=1)
for(i in seq(from=0, to=0.16, len=35)){
    points(ray(i), type='l', lwd=0.1)
}
```

```{r definefinaldeflection}
final_deflection <- function(s){
    jj <- ray(s, give=TRUE)
    jj[nrow(jj),3]
}

final_deflection(0.3)
final_deflection(0.31)
head(ray(0.1,give=TRUE))
tail(ray(0.1,give=TRUE))
```

```{r finaldefrange, cache=TRUE}
s <- seq(from=0, to=0.25, len=10)
fa <- sapply(s, final_deflection)
```

```{r}
plot(s, fa, type="b")
```

Find the Cartesian ray:

```{r minimizefinaldef, cache=TRUE}
optimize(final_deflection, interval = c(0.1,0.15))
```


# Catenary

Now 

\begin{equation}
F = \sqrt{1+y'^2}(y + \Lambda)\qquad G =  \sqrt{1+y'^2}(\phi + \lambda)
\end{equation}

where $\phi=\phi(x,y)$ is a small perturbation.

\begin{eqnarray}
\el{F} &=& \frac{d}{dx}\left[\frac{(y+\Lambda)y'}{\sqrt{1+y'^2}}\right] - \sqrt{1+y'^2}\nonumber\\
       &=& \frac{y''(y+\Lambda)}{(1+y'^2)^{3/2}} + \frac{y'^2}{\sqrt{1+y'^2}} - \sqrt{1+y'^2}\nonumber\\
       &=& \frac{y''(y+\Lambda)}{(1+y'^2)^{3/2}} - \frac{1}{\sqrt{1+y'^2}}\label{elF}
\end{eqnarray}

and 

\begin{eqnarray}
\el{G} &=& \frac{d}{dx}\di{G}{y'} - \phi_y\sqrt{1+y'^2}\nonumber\\
       &=& \frac{d}{dx}\left[\frac{y'(\phi+\lambda)}{\sqrt{1+y'^2}}\right] - \phi_y\sqrt{1+y'^2}\nonumber\\
       &=& \frac{y''(\phi+\lambda)}{(1+y'^2)^{3/2}} + \frac{y'(\phi_x + y'\phi_y)}{\sqrt{1+y'^2}} - \phi_y\sqrt{1+y'^2}\nonumber\\
       &=& \frac{y''(\phi+\lambda)}{(1+y'^2)^{3/2}} + \frac{1}{\sqrt{1+y'^2}}\left(y'(\phi_x + y'\phi_y) - \phi_y-\phi_yy^2\right)\nonumber\\
       &=& \frac{y''(\phi+\lambda)}{(1+y'^2)^{3/2}} + \frac{y'\phi_x  -\phi_y}{\sqrt{1+y'^2}}\label{elG}
\end{eqnarray}


Restating for convenience:

\begin{equation}
\el{G} + \left(z\di{}{y} + z'\di{}{y'} + z''\di{}{y''}\right)\label{feqnagain}
\left(\el{F}\right)=0
\end{equation}

where $\el{F} = \frac{(y+\Lambda)y''}{(1+y'^2)^{3/2}} -
\frac{1}{\sqrt{1+y'^2}}$ and $\el{G} =
\frac{y''(\phi+\lambda)}{(1+y'^2)^{3/2}} + \frac{y'\phi_x
-\phi_y}{\sqrt{1+y'^2}}$.  The first term of equation \ref{feqnagain}
is given by equation \ref{elG}.  For the second:

\begin{eqnarray}
 &&  \left(z\di{}{y} + z'\di{}{y'} + z''\di{}{y''}\right)
     \left(\el{F}\right)\nonumber\\
&=&  \left(z\di{}{y} + z'\di{}{y'} + z''\di{}{y''}\right)
     \left(\frac{y''(y+\Lambda)}{(1+y'^2)^{3/2}} - \frac{1}{\sqrt{1+y'^2}}\right)\nonumber\\
&=& 
  z  \left(\frac{y''}{(1+y'^2)^{3/2}}\right)
+ z' \left(\frac{-3(y+\Lambda)y'y''}{(1+y'^2)^{5/2}} - \frac{y'}{(1+y'^2)^{3/2}}\right)
+ z''\left(\frac{y+\Lambda}{(1+y'^2)^{3/2}}\right)
\label{zyzdddyzdddydd}
\end{eqnarray}


So altogether equation \ref{feqnagain} becomes

\begin{eqnarray}
\underbrace{ \frac{y''(\phi+\lambda)}{(1+y'^2)^{3/2}} + \frac{y'\phi_x
-\phi_y}{\sqrt{1+y'^2}}}_{\mbox{equation \ref{elG}}}
+ \underbrace{
  z  \frac{y''}{(1+y'^2)^{3/2}}
+ z' \left(\frac{-3(y+\Lambda)y'y''}{(1+y'^2)^{5/2}} - \frac{y'}{(1+y'^2)^{3/2}}\right)
+ z''\left(\frac{y+\Lambda}{(1+y'^2)^{3/2}}\right)
}_{\mbox{equation \ref{zyzdddyzdddydd}}}
=0\nonumber\\
\frac{y''(\phi+\lambda)}{(1+y'^2)^{3/2}} + \frac{y'\phi_x-\phi_y}{\sqrt{1+y'^2}} +
  z  \frac{y''}{(1+y'^2)^{3/2}}
- z' \left(\frac{3(y+\Lambda)y'y''}{(1+y'^2)^{5/2}} + \frac{y'}{(1+y'^2)^{3/2}}\right)
+ z''\left(\frac{y+\Lambda}{(1+y'^2)^{3/2}}\right) = 0
\end{eqnarray}

Now for the unperturbed catenary we use $y=\cosh x$ [with $\Lambda=0$]
giving $y'=\sinh x$ and $y''=\cosh x$.  In particular we note
$\sqrt{1+y'^2}=\cosh x$.  Thus:


\begin{eqnarray}
\frac{y''(\phi+\lambda)}{(1+y'^2)^{3/2}} + \frac{y'\phi_x-\phi_y}{\sqrt{1+y'^2}}
+ z  \frac{y''}{(1+y'^2)^{3/2}}
- z' \left(\frac{3yy'y''}{(1+y'^2)^{5/2}} + \frac{y'}{(1+y'^2)^{3/2}}\right)
+ z''\left(\frac{y}{(1+y'^2)^{3/2}}\right) = 0\nonumber\\
\frac{\cosh x(\phi+\lambda)}{\cosh^3 x} + \frac{\phi_x\sinh x-\phi_y}{\cosh x}
+ z  \frac{\cosh x}{\cosh^3 x}
- z' \left(\frac{3\cosh^2 x\sinh x}{\cosh^5 x} + \frac{\sinh x}{\cosh^3 x}\right)
+ z''\left(\frac{\cosh x}{\cosh^3x}\right) = 0\nonumber\\
\frac{\phi+\lambda}{\cosh^2 x} + \frac{\phi_x\sinh x-\phi_y}{\cosh x}
+          \frac{z}{\cosh^2 x}
- z' \left(\frac{3\sinh x}{\cosh^3 x} + \frac{\sinh x}{\cosh^3 x}\right)
+          \frac{z''}{\cosh^2x} = 0\nonumber\\
  \frac{\phi+\lambda}{\cosh^2 x} + \frac{\phi_x\sinh x-\phi_y}{\cosh x}
+ \frac{z}{\cosh^2 x}
- \frac{4z'\sinh x}{\cosh^3 x}
+ \frac{z''}{\cosh^2x} = 0\nonumber\\
  \phi + \lambda + \phi_x\sinh x\cosh x - \phi_y\cosh x + z - 4z'\tanh x + z'' = 0
\end{eqnarray}


\begin{equation}
z'' = 4z'\tanh x -z  -(\phi + \lambda + \phi_x\sinh x\cosh x - \phi_y\cosh x)\label{zdashdash}
\end{equation}

Above we understand that $\phi=\phi(x,y)=\phi(x,\cosh x)$.

## Numerical work for the catenary


We solve equation \ref{zdashdash} using the same numerical method as
for the straight line.

```{r catenaryODEdefiner}
catenary <- function(t, state, parameters) {
    with(as.list(c(state, parameters)), {
        dZ <- Zdash

        x <- t
        y <- cosh(x)
        
        phi <- x
        phi_y  <- 0
        phi_x <- 1
        

        dZdash <- 4*Zdash*tanh(x) - Z -(phi + lambda + phi_x*sinh(x)*cosh(x) - phi_y*cosh(x)) # equation 24

        dextra_length <- Zdash * tanh(x)
        dextra_length2 <- -Z/cosh(x)^2
        list(c(dZ, dZdash, dextra_length, dextra_length2))
    })
}
```


```{r solvecatenaryODE, cache=TRUE}

parameters <- c(lambda = 0)
state      <- c(Z = 0, Zdash = 0, extra_length = 0, extra_length2 = 0)  # initial conditions
times      <- seq(-2, 2, by = 0.1)

out <- ode(y = state, times = times, func = catenary, parms = parameters)
```

```{r showcatenaryODE}
head(out)
tail(out)
x <- out[,1]
y <- out[,2]
plot(x,y,type='l')
grid()
```

```{r definecatenarybadness}
badness <- function(unknown, give=FALSE){
    lambda        <- unknown[1]
    Zdash_initial <- unknown[2]
    times <- seq(from = -2, to = 2, by = 1)
    parameters <- c(lambda = lambda)
    state <-  # initial conditions
        c(Z = 0, Zdash = Zdash_initial, extra_length=0, extra_length2=0) 
    out <- ode(y = state, times = times, func = catenary, parms = parameters)
    final_disp <- out[5,2]
    extra_length <- out[5,4]
    badness <- final_disp^2 + extra_length^2
    badness <- abs(final_disp) + abs(extra_length)
    if(give){
        return(out)
    } else {
        return(badness)
    } 
}
```


```{r}
badness(c(5, -9))
badness(c(6, -9))
```


```{r usecatenarybadness, cache=TRUE}
(jj<- optim(c(0,0),badness, hessian = TRUE))
minbad <- jj$par

parameters <- c(lambda = minbad[1])
state <- c(Z = 0, Zdash = minbad[2], extra_length = 0, extra_length2 = 0)
times <- seq(-2, 2, by = 0.1)
# initial conditions
out <- ode(y = state, times = times, func = catenary, parms = parameters)
head(out)
tail(out)
```

```{r}
jj$hessian
eigen(jj$hessian)
```

```{r usef, cache=TRUE}
v1 <- seq(from = -0.1, to = 0.1, len=15)
v2 <- seq(from = -12.1, to = -11.9, len=65)
M <- as.matrix(expand.grid(v1, v2))
head(M)
bad <- sapply(seq_len(nrow(M)),function(i){badness(unname(M[i,]))})
```

```{r contbad, cache=FALSE}
M <- matrix(bad, length(v1), length(v2))
M <- pmin(M, Inf)
contour(v1, v2, M)
filled.contour(M)
```

It seems that the contours of badness (on the lambda-gradient plane)
are elliptical but somewhat elongated.  It is further clear that the
optimal value of lambda is zero; what is the optimal value of the
initial gradient at this point?


```{r conditionalbadness, cache=TRUE}
bad2 <- function(lambda, give=FALSE){
    out <- optimize(
        f = function(initial_gradient){badness(c(lambda, initial_gradient))},
        interval = c(-20, 20)
    )
    if(!give){out <- out$minimum}
    return(out)
}

(jj <- bad2(0, give=T))
optimal_initial_gradient <- jj$minimum
```


```{r lambad, cache=TRUE}
lam <- seq(from = -0.05, to = 0.05, by=0.01)
bad <- sapply(lam, function(l){bad2(l, give=TRUE)$objective})
```

```{r plotlambad}
plot(lam, bad, type = 'both')
```


```{r}
parameters <- c(lambda = 0)
state <- c(Z = 0, Zdash = optimal_initial_gradient, extra_length = 0, extra_length2 = 0)
times <- seq(-2, 2, by = 0.1)
# initial conditions
cat_perturbation <- ode(y = state, times = times, func = catenary, parms = parameters)
head(cat_perturbation)
tail(cat_perturbation)
```

```{r}
x  <- cat_perturbation[,1]
dy <- cat_perturbation[,2]
plot(x,dy)

plot(x, cosh(x), type='l')
points(x, cosh(x) + dy*0.1,col='red',type='l')
```


# Now a perturbed catenary on a rotating spaceship

Imagine you are in a spaceship that is rotating in order to simulate
Earth's gravitational field.  You hold a chain in both hands and it
forms a slightly perturbed catenary; in fact it minimizes $F+\epsilon
G$ where

\begin{equation}
F = \sqrt{1+y'^2}(y+\Lambda),\qquad
G = \sqrt{1+y'^2}(y^2+\lambda)
\end{equation}

Then

\begin{eqnarray}
\el{F} 
&=& \frac{d}{dx}\left(\frac{y'(y+\Lambda)}{\sqrt{1+y'^2}}\right) - \sqrt{1+y'^2}\nonumber\\
&=& \frac{(y+\Lambda)y''}{(1+y'^2)^{3/2}} + \frac{y'^2}{\sqrt{1+y'^2}} - \sqrt{1+y'^2}\nonumber\\
&=& \frac{(y+\Lambda)y''}{(1+y'^2)^{3/2}} -\frac{1}{\sqrt{1+y'^2}}\\
\end{eqnarray}

and 

\begin{eqnarray}
\el{G} 
&=& \frac{d}{dx}\left(\frac{y'(y^2+\Lambda)}{\sqrt{1+y'^2}}\right)  - 2y\sqrt{1+y'^2}\nonumber\\
&=& \frac{(y^2+\Lambda)y''}{(1+y'^2)^{3/2}} + \frac{2yy'^2}{\sqrt{1+y'^2}}- 2y\sqrt{1+y'^2}\nonumber\\
&=& \frac{(y^2+\Lambda)y''}{(1+y'^2)^{3/2}} - \frac{2y}{\sqrt{1+y'^2}}\nonumber\\
\end{eqnarray}

\begin{eqnarray}
\el{G} + \left(z\di{}{y} + z'\di{}{y'} + z''\di{}{y''}\right)\left(\el{F}\right) &=& 0\nonumber\\
\frac{(y^2+\Lambda)y''}{(1+y'^2)^{3/2}} - \frac{2y}{\sqrt{1+y'^2}}
+ \left(z\di{}{y} + z'\di{}{y'} +
z''\di{}{y''}\right)\left(\frac{(y^2+\Lambda)y''}{(1+y'^2)^{3/2}} -
\frac{2y}{\sqrt{1+y'^2}}\right) &=& 0\nonumber\\
\frac{(y^2+\Lambda)y''}{(1+y'^2)^{3/2}} - \frac{2y}{\sqrt{1+y'^2}}
+ z  \left(\frac{y}{(1+y'^2)^{3/2}} - \frac{2}{\sqrt{1+y'^2}}\right)\nonumber\\
+ z' \left(\frac{-3(y^2+\Lambda)y'y''}{(1+y'^2)^{5/2}} + \frac{2yy'}{(1+y'^2)^{3/2}}\right)
+ z''\frac{y^2+\Lambda}{(1+y'^2)^{3/2}} &=& 0
\end{eqnarray}

again, we set $\Lambda=0$ and with boundary conditions we get
$y'=y''=\sqrt{1+y'^2}=\cosh x$, $y'=\sinh x$:

\begin{eqnarray}
\frac{(y^2+\Lambda)y''}{(1+y'^2)^{3/2}} - \frac{2y}{\sqrt{1+y'^2}}
+ z  \left(\frac{y}{(1+y'^2)^{3/2}} - \frac{2}{\sqrt{1+y'^2}}\right)\nonumber\\
+ z' \left(\frac{-3(y^2+\Lambda)y'y''}{(1+y'^2)^{5/2}} + \frac{2yy'}{(1+y'^2)^{3/2}}\right)
+ z''\frac{y^2+\Lambda}{(1+y'^2)^{3/2}} &=& 0\nonumber\\
\frac{y^2y''}{(1+y'^2)^{3/2}} - \frac{2y}{\sqrt{1+y'^2}}
+ z  \left(\frac{y}{(1+y'^2)^{3/2}} - \frac{2}{\sqrt{1+y'^2}}\right)\nonumber\\
+ z' \left(\frac{-3y^2y'y''}{(1+y'^2)^{5/2}} + \frac{2yy'}{(1+y'^2)^{3/2}}\right)
+ z''\frac{y^2}{(1+y'^2)^{3/2}} &=& 0\nonumber\\
\frac{\cosh^3 x}{\cosh^3 x} - \frac{2\cosh x}{\cosh x}
+ z  \left(\frac{\cosh x}{\cosh^3 x} - \frac{2}{\cosh x}\right)\nonumber\\
+ z' \left(\frac{-3\cosh^3 x\sinh x}{\cosh^5x} + \frac{2\sinh x\cosh x}{\cosh^3x}\right)
+ z''\frac{\cosh^2x}{\cosh^3x} &=& 0\nonumber\\
-1 +z\left(\frac{1}{\cosh^2x}-\frac{2}{\cosh x}\right) - z'\left(\frac{\sinh x}{\cosh x}\right)
+z''\left(\frac{1}{\cosh x}\right) &=& 0\nonumber\\
+z\left(\frac{1}{\cosh^2x}-\frac{2}{\cosh x}\right) - z'\left(\frac{\sinh x}{\cosh x}\right)
+z''\left(\frac{1}{\cosh x}\right) &=& 1\nonumber\\
+z\left(\cosh^{-1}x - 2\right) - z'\tanh x
+z'' &=& 1\nonumber\\
z\left(2-\cosh^{-1}x\right) + z'\tanh x  +1 &=& z''
\end{eqnarray}


