---
title: "Perturbing the catenary: a first-order analysis"
author: "Robin Hankin"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The catenary.

\newcommand{\di}[2]{\frac{\partial{#1}}{\partial{#2}}}
\newcommand{\cx}{\cosh{\left(\frac{x-x_0}{a}\right)}}
\newcommand{\sx}{\sinh{\left(\frac{x-x_0}{a}\right)}}
\newcommand{\tx}{\tanh{\left(\frac{x-x_0}{a}\right)}}
\newcommand{\sech}{\operatorname{sech}}

We start with the Euler-Lagrange equation

\begin{equation}
\frac{d}{dx}\di{F}{y'} -\di{F}{y}=0
\end{equation}

and if $\di{F}{x}=0$ we have the Beltrami equation

\begin{equation}
F-y'\di{F}{y'}=C
\end{equation}

For a catenary we have

\begin{equation}
F=\sqrt{1+(y')^2}(y + \lambda)
\end{equation}

so Beltrami applies giving

\begin{equation}
\sqrt{1+y'^2}(y + \lambda)-\frac{y'^2(y+\lambda)}{\sqrt{1+y'^2}}=C
\end{equation}

Simplifying

\begin{equation}
(y+\lambda)=C\sqrt{1+y'^2}
\end{equation}

or

\begin{equation}
dx=\frac{dy}{\sqrt{\left(\frac{y+\lambda}{C}\right)^2-1}}
\end{equation}

Substituting $u=(y+\lambda)/C$ so $du=dy/C$ we have

\begin{equation}
x=\int\frac{C\,du}{\sqrt{u^2-1}} =
C\operatorname{acosh}\left(\frac{y+\lambda}{C}\right)+K
\end{equation}

giving, in more conventional notation,

\begin{equation}
y = y_0 + a\operatorname{cosh}\left(\frac{x-x_0}{a}\right)
\end{equation}

Here for simplicity we use $x_0=y_0=0$, $a=1$.


# Perturbing the catenary.

The first-order corrective term for EL is:

\begin{equation}\label{EL1}
\frac{d}{dx}\left[\di{}{y'}\left(
z\di{F}{y} + z'\di{F}{y'} + G\right)\right] =
\di{}{y}\left(z\di{F}{y} + z'\di{F}{y'} + G\right)
\end{equation}


Now we consider a flexible chain minimising its potential energy in a
slightly perturbed potential field $y + \epsilon\phi(x,y)$ where
$\epsilon$ is sufficiently small to neglect $\epsilon^2$ and higher.
Here we have $F=\sqrt{1+y'^2}(y+\lambda)$ and
$G=\sqrt{1+y'^2}\phi(x,y)$.

Before we apply it to the catenary we need to consider the form of
$G$, which will include a constraint.  We have
$F=\sqrt{1+y'^2}(y+\lambda)$ giving
$y(x)=a\cosh\left(\frac{x-x_0}{a}\right)$.  At first glance, the
perturbation $\epsilon G$ is just $\epsilon\phi(x,y)\sqrt{1+y'^2}$.
But there is a constraint here.  The overall length should not be
altered.  As above, we have

\begin{eqnarray}
\int_{x=-a}^{x=+a}\sqrt{1+y'^2}dx &\longrightarrow& \int_{x=-a}^{x=+a}\sqrt{1+(y+\epsilon z)'^2}dx\\
&=& \int_{x=-a}^{x=+a}\sqrt{1+y'^2 +2\epsilon y'z'}dx \\
&=& \int_{x=-a}^{x=+a}\left(\sqrt{1+y'^2} + \epsilon\frac{y'z'}{\sqrt{1+y'^2}}\right)dx
\end{eqnarray}


So the constraint should be that 

\begin{equation}
\int_{x=-a}^{x=+a}\frac{y'z'\,dx}{\sqrt{1+y'^2}}=0
\end{equation}

Reducing to 

\begin{equation}
\int_{x=-a}^{x=+a}z'(x)\tanh(x)\,dx=0
\end{equation}

Using the product rule we get

\begin{equation}
\int_{x=-a}^{x=+a}z(x)\sech^2 x\,dx=0
\end{equation}


So the appropriate constraint on G would be 

\begin{equation}
G = \phi(x,y)\sqrt{1+y'^2} + \mu z'(x)\frac{y'}{\sqrt{1+y'^2}}
\end{equation}

where $\mu$ is an undetermined multiplier.  BUT THIS IS WRONG.  G is
not being optimized (actually $F+\epsilon G$ is being optimized).  For
convenience here is equation \ref{EL1}:

\begin{equation}
\frac{d}{dx}\left[\di{}{y'}\left(
z\di{F}{y} + z'\di{F}{y'} + G\right)\right] =
\di{}{y}\left(z\di{F}{y} + z'\di{F}{y'} + G\right)\nonumber
\end{equation}


\begin{equation}
F = (y+\lambda)\sqrt{1+y'^2}+ \frac{\mu y'z'}{\sqrt{1+y'^2}},\qquad 
G = \phi(x,y)\sqrt{1+y'^2}
\end{equation}


\begin{equation}
\frac{d}{dx}\di{H}{y'}=\di{H}{y},\qquad
H=z\di{F}{y} + z'\di{F}{y'} + G
\end{equation}

We find that 

\begin{eqnarray}
H &=& z\di{F}{y} + z'\di{F}{y'} + G \nonumber\\
  &=& z\underbrace{\sqrt{1+y'^2}}_{\partial F/\partial y} + 
      z'
	  \underbrace{\left(
	     \frac{y'(y+\lambda)}{\sqrt{1+y'^2}} + \frac{\mu}{(1+y'^2)^{3/2}}
	     \right)
	  }_{
	     \partial F/\partial y'
	  } + 
      \underbrace{\phi(x,y)\sqrt{1+y'^2}}_{G}\nonumber\\
  &=& (z+\phi(x,y))\sqrt{1+y'^2}  + \frac{y'(y+\lambda)z'}{\sqrt{1+y'^2}} + \frac{\mu z'}{(1+y'^2)^{3/2}}
\end{eqnarray}

We have

\begin{equation}
\di{H}{y'} = 
  \frac{(z+\phi)y'}{\sqrt{1+y'^2}}
+ \frac{(y+\lambda)z'}{(1+y'^2)^{3/2}}
- \frac{3\mu y'z'}{(1+y'^2)^{5/2}}
\end{equation}

\begin{equation}
\di{H}{y} = 
    \phi_y\sqrt{1+y'^2}
 +  \frac{y'z'}{\sqrt{1+y'^2}}
 \end{equation}

\begin{eqnarray}
\frac{d}{dx}\di{H}{y'} &=& \di{H}{y}\nonumber\\
\frac{d}{dx}
\left[
  \frac{(z+\phi)y'}{\sqrt{1+y'^2}}
+ \frac{(y+\lambda)z'}{(1+y'^2)^{3/2}}
- \frac{3\mu y'z'}{(1+y'^2)^{5/2}}
\right]
&=& 
    \phi_y\sqrt{1+y'^2}
 +  \frac{y'z'}{\sqrt{1+y'^2}}
\end{eqnarray}

Now we remind ourselves that $y=y_0+\cx$ and $y'=\sx$, so
$\sqrt{1+y'^2}=\cx$ and set $x_0=y_0=0$, $a=1$, so $y=\cosh x$ and
$y'=\sinh x$:

\begin{eqnarray}
\frac{d}{dx}
\left[
\frac{z'(\lambda+\mu+\cosh x)}{\cosh^3 x} +
(z+\phi(x,y))\tanh x
\right] &=&
\phi_y(x,y)\cosh x + z'\tanh x
\end{eqnarray}


\begin{eqnarray}
\frac{d}{dx}
\left[
z'(\lambda\sech^3 x +\sech^2 x) + 
(z+\phi(x,y))\tanh x
\right] &=&
\phi_y(x,y)\cosh x + z'\tanh x
\end{eqnarray}


\begin{eqnarray}
LHS &=&    z''(\lambda\sech^3 x +\sech^2 x) - z'(3\lambda\sech^3x\tanh x + 2\sech^2\tanh x)\nonumber\\
    &{}& +(z'+\phi_x + \phi_y\sinh x)\tanh x + (z+\phi)\sech^2 x\\ 
    &=&   z''(\lambda\sech^3 x +\sech^2 x)-  z'(3\lambda\sech^3x\tanh x + 2\sech^2\tanh x-\tanh x)\nonumber\\
    &{}& + z\sech^2 x + (\phi_x + \phi_y\sinh x)\tanh x + \textcolor{red}{\phi\sech^2 x}\\
    &=&   z''(\lambda\sech^3 x +\sech^2 x)-  z'(3\lambda\sech^3x\tanh x + 2\sech^2\tanh x-\tanh x)\nonumber\\
    &{}&  + (\phi_x + \phi_y\sinh x)\tanh x + (z+\textcolor{red}{\phi})\sech^2 x
	\end{eqnarray}

\begin{equation}
RHS = \phi_y\cosh x + z'\tanh x
\end{equation}

So $LHS=RHS$ becomes

\begin{eqnarray}
  &{}& z''(\lambda\sech^3 x +\sech^2 x) - z'(3\lambda\sech^3x\tanh x + 2\sech^2\tanh x-\tanh x)\nonumber\\
    &{}&  + (\phi_x + \phi_y\sinh x)\tanh x + (z+\textcolor{red}{\phi})\sech^2 x = \phi_y\cosh x + z'\tanh x
\end{eqnarray}

\begin{eqnarray}
  &{}& z''(\lambda\sech^3 x +\sech^2 x) - z'(3\lambda\sech^3x\tanh x + 2\sech^2\tanh x)\nonumber\\
    &{}&  + (\phi_x + \phi_y\sinh x)\tanh x + (z+\textcolor{red}{\phi})\sech^2 x
	= \phi_y\cosh x 
\end{eqnarray}

\begin{eqnarray}
   z'' = 
\frac{
     z'\cdot (3\lambda\sech^3x\tanh x + 2\sech^2\tanh x)
  -  z \cdot \sech^2x
  + \phi_y\cosh x - (\phi_x + \phi_y\sinh x)\tanh x - \textcolor{red}{\phi}\sech^2 x
}{
\lambda\sech^3 x +\sech^2 x
}
\end{eqnarray}




# Numerical work


```{r odesolver}
library("deSolve") 
sech <- function(x){1/cosh(x)}
Lorenz <- function(t, state, parameters) {
    with(as.list(c(state, parameters)), {
        dZ <- Zdash

        phi <- t-1
        phi_y  <- 0
        phi_x <- 1
        numerator <- (
            +Zdash * (3*lambda * sech(t)^3 * tanh(t) + 2*sech(t)^2 * tanh(t))
            -Z     * sech(t)^2
            + phi_y * cosh(t)
            -(phi_x + phi_y*sinh(t))*tanh(t)
            -phi*sech(t)^2
        )

        denominator <- lambda *sech(t)^3  + sech(t)^2

        dZdash <- numerator/denominator
        dextra_length <- Zdash*tanh(t)
        dextra_length2 <- -Z*sech(t)^2
        list(c(dZ, dZdash, dextra_length, dextra_length2))
    })
}

parameters <- c(lambda = 0)
state      <- c(Z = 0, Zdash = 0, extra_length = 0, extra_length2 = 0)  # initial conditions
times      <- seq(-2, 2, by = 0.05)

out <- ode(y = state, times = times, func = Lorenz, parms = parameters)
```

```{r showode}
head(out)
tail(out)
x <- out[,1]
y <- out[,2]
plot(x,y,type='l')
grid()
```


```{r defodefunc}
f <- function(unknown, give=FALSE){
    lambda        <- unknown[1]
    Zdash_initial <- unknown[2]
    times <- seq(from = -2, to = 2, by = 1)
    parameters <- c(lambda = lambda)
    state <-  # initial conditions
        c(Z = 0, Zdash = Zdash_initial, extra_length=0, extra_length2=0) 
    out <- ode(y = state, times = times, func = Lorenz, parms = parameters)
    final_disp <- out[5,2]
    extra_length <- out[5,4]
    badness <- final_disp^2 + extra_length^2
    if(give){
        return(out)
    } else {
        return(badness)
    } 
}
```

```{r}
f(c(2, 1.0))
f(c(2, 1.1))
```
   
   
```{r cache=TRUE}
jj <- optim(c(0,0),f)
jj

```

```{r usef, cache=TRUE}
v1 <- seq(from = 0, to = 3, by = 0.1)
v2 <- seq(from = -5, to = -2, by = 0.1)
M <- as.matrix(expand.grid(v1,v2))
head(M)
bad <- sapply(seq_len(nrow(M)),function(i){f(unname(M[i,]))})
```

```{r contbad}
M <- matrix(bad,length(v1),length(v2))
contour(v1, v2, M, levels=0:10)
filled.contour(M)
```

next chunk

	
```{r}
jj$par
parameters <- c(lambda = jj$par[1])
state      <- c(Z = 0, Zdash = jj$par[2], extra_length=0, extra_length2=0)  # initial conditions
times      <- seq(-2, 2, by = 0.01)

out <- ode(y = state, times = times, func = Lorenz, parms = parameters)
head(out)
tail(out)
x  <- out[,1]
z <- out[,2]
y0 <- cosh(x)

plot(x, z,type="l")
plot(x, y0,type="l")
points(x, y0+z/10,type="l",col="red")
```


OK we have two points that are not the same but both satisfy the
boundary conditions and indeed the length constraint.
n


```{r}
p1 <- c(4,-3.409)
p2 <- c(3,-3.999)
```

```{r}
show <- function(p){
    parameters <- c(lambda = p[1])
state      <- c(Z = 0, Zdash = p[2], extra_length=0, extra_length2=0)  # initial conditions
times      <- seq(-2, 2, by = 0.01)
ode(y = state, times = times, func = Lorenz, parms = parameters)
}


jj1 <- show(p1)
jj2 <- show(p2)

plot(times,jj2[,2])
points(times,jj1[,2])


plot(times,jj1[,2]/jj2[,2])
```


```{r}
jj3 <- show(c(0, -8.7563))
plot(times,jj3[,2])
```

```{r}
parameters <- c(lambda = 0)
state      <- c(Z = 0, Zdash = -8.7563, extra_length=0, extra_length2=0)  # initial conditions
times      <- seq(-2, 2, by = 0.01)

out <- ode(y = state, times = times, func = Lorenz, parms = parameters)
head(out)
tail(out)
x  <- out[,1]
z <- out[,2]
y0 <- cosh(x)

plot(x, z,type="l")
plot(x, y0,type="l")
points(x, y0+z/5,type="l",col="red")
```


# Exact solutions

Try to solve the exact EL equation in a potential of $y+ax$
[notionally $|a|\ll 1$]


\begin{equation}
F=(y+ax+\lambda)\sqrt{1+y'^2}
\end{equation}

\begin{equation}
\frac{d}{dx}\di{F}{y'}=\di{F}{y}
\end{equation}


\begin{equation}
\frac{d}{dx}\left[\frac{(y+ax+\lambda)y'}{\sqrt{1+y'^2}}\right]=\sqrt{1+y'^2}
\end{equation}

\begin{equation}
\frac{(y'+a)y'}{\sqrt{1+y'^2}} + 
\frac{y''(y+ax+\lambda)}{(1+y'^2)^{3/2}}=\sqrt{1+y'^2}
\end{equation}

\begin{equation}
(y'^2+ay')(1+y^2) + y''(y+ax+\lambda) = (1+y'^2)^2
\end{equation}

\begin{equation}
y''(y+ax+\lambda) = (1+y'^2)(1-ay'^2)
\end{equation}

\begin{equation}
y'' = \frac{(1+y'^2)(1-ay'^2)}{y+ax+\lambda}
\end{equation}

# Exact solutions (numerical work)

Here we consider an inextensible string in a potential
$\phi(x,y)=ax+y$.  So $a=0$ is vertical gravitational field.

```{r}
cat <- function(t, state, parameters) {
    with(as.list(c(state, parameters)), {
        dZ <- Zdash
        numerator <- (1+Zdash^2) * (1-a*Zdash^2)
        denominator <-  Z + a*t + lambda
        dZdash <- numerator/denominator
        list(c(dZ, dZdash))
    })
}

parameters <- c(lambda = 1.5, a = 0.2)
state      <- c(Z = 0, Zdash = -1)  # initial conditions
times      <- seq(-2, 1, by = 0.01)

out <- ode(y = state, times = times, func = cat, parms = parameters)
```

```{r showex}
head(out)
tail(out)
x <- out[,1]
y <- out[,2]
plot(x,y,asp=1,type='l')
grid()
abline(5.0,1/0.2,lwd=0.2)
abline(5.2,1/0.2,lwd=0.1)
abline(0.0,-0.2,lwd=0.2)
abline(0.2,-0.2,lwd=0.1)
```


