---
title: "Perturbing the catenary: a first-order analysis"
author: "Robin Hankin"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The catenary.

\newcommand{\di}[2]{\frac{\partial{#1}}{\partial{#2}}}
\newcommand{\cx}{\cosh{\left(\frac{x-x_0}{a}\right)}}
\newcommand{\sx}{\sinh{\left(\frac{x-x_0}{a}\right)}}
\newcommand{\tx}{\tanh{\left(\frac{x-x_0}{a}\right)}}
\newcommand{\sech}{\operatorname{sech}}

We start with the Euler-Lagrange equation

\begin{equation}
\frac{d}{dx}\di{F}{y'} -\di{F}{y}=0
\end{equation}

and if $\di{F}{x}=0$ we have the Beltrami equation

\begin{equation}
F-y'\di{F}{y'}=C
\end{equation}

For a catenary we have

\begin{equation}
F=\sqrt{1+(y')^2}(y + \lambda)
\end{equation}

so Beltrami applies giving

\begin{equation}
\sqrt{1+y'^2}(y + \lambda)-\frac{y'^2(y'+\lambda)}{\sqrt{1+y'^2}}=C
\end{equation}

Simplifying

\begin{equation}
(y+\lambda)=C\sqrt{1+y'^2}
\end{equation}

or

\begin{equation}
dx=\frac{dy}{\sqrt{\left(\frac{y+\lambda}{C}\right)^2-1}}
\end{equation}

Substituting $u=(y+\lambda)/C$ so $du=dy/C$ we have

\begin{equation}
x=\int\frac{C\,du}{\sqrt{u^2-1}} =
C\operatorname{acosh}\left(\frac{y+\lambda}{C}\right)+K
\end{equation}

giving, in more conventional notation,

\begin{equation}
y = y_0 + a\operatorname{cosh}\left(\frac{x-x_0}{a}\right)
\end{equation}

Here for simplicity we use $x_0=y_0=0$, $a=1$.


# Perturbing the catenary.

The first-order corrective term for EL is:

\begin{equation}\label{EL1}
\frac{d}{dx}\left[\di{}{y'}\left(
z\di{F}{y} + z'\di{F}{y'} + G\right)\right] =
\di{G}{y}
\end{equation}


Now we consider a flexible chain minimising its potential energy in a
slightly perturbed potential field $y + \epsilon\phi(x,y)$ where
$\epsilon$ is sufficiently small to neglect $\epsilon^2$ and higher.
Here we have $F=\sqrt{1+y'^2}(y+\lambda)$ and
$G=\sqrt{1+y'^2}\phi(x,y)$.

Before we apply it to the catenary we need to consider the form of
$G$, which will include a constraint.  We have
$F=\sqrt{1+y'^2}(y+\lambda)$ giving
$y(x)=a\cosh\left(\frac{x-x_0}{a}\right)$.  At first glance, the
perturbation $\epsilon G$ is just $\epsilon\phi(x,y)\sqrt{1+y'^2}$.
But there is a constraint here.  The overall length should not be
altered.  As above, we have

\begin{eqnarray}
\int_{x=-a}^{x=+a}\sqrt{1+y'^2}dx &\longrightarrow& \int_{x=-a}^{x=+a}\sqrt{1+(y+\epsilon z)'^2}dx\\
&=& \int_{x=-a}^{x=+a}\sqrt{1+y'^2 +2\epsilon y'z'}dx \\
&=& \int_{x=-a}^{x=+a}\left(\sqrt{1+y'^2} + \epsilon\frac{y'z'}{\sqrt{1+y'^2}}\right)dx
\end{eqnarray}


So the constraint should be that 

\begin{equation}
\int_{x=-a}^{x=+a}\frac{y'z'\,dx}{\sqrt{1+y'^2}}=0
\end{equation}

So the appropriate constraint on G would be 

\begin{equation}
G = \phi(x,y)\sqrt{1+y'^2} + \mu z'(x)\frac{y'}{\sqrt{1+y'^2}}
\end{equation}

where $\mu$ is an undetermined multiplier.  For convenience here is
equation \ref{EL1}:

\begin{equation}\nonumber
\frac{d}{dx}\left[\di{}{y'}\left(
z\di{F}{y} + z'\di{F}{y'} + G\right)\right] =
\di{G}{y}
\end{equation}

\begin{eqnarray}
\frac{d}{dx}\left[\di{}{y'}\left(
z\di{F}{y} + z'\di{F}{y'}
+ \underbrace{\phi(x,y)\sqrt{1+y'^2} + \mu z'(x)\frac{y'}{\sqrt{1+y'^2}}}_{G(x,y,y')}
\right)\right] &=& \di{\phi}{y}\sqrt{1+y'^2}
\end{eqnarray}

(above we have substituted for $G$).  But now we know that
$F=\sqrt{1+y'^2}(y+\lambda)$ so we now have

\begin{eqnarray}
\di{F}{y} &=&\sqrt{1+y'^2}=\cosh x\\
\di{F}{y'} &=&
\frac{y'(y+\lambda)}{\sqrt{1+y'^2}}=\tanh x\left(\cosh x+\lambda\right)\\
\frac{\partial^2F}{\partial y\partial y'} &=& \frac{y'}{\sqrt{1+y'^2}}=\tanh x\\
\frac{\partial^2F}{\partial y'\partial y'} &=&
\frac{y+\lambda}{(1+y'^2)^{3/2}} = \frac{\cosh x + \lambda}{\cosh^3 x}
\end{eqnarray}

Thus

\begin{eqnarray}
\frac{d}{dx}\left[
z\tanh x + z'\frac{\cosh x+\lambda}{\cosh^3x}
+ \phi(x,y)\frac{y'}{\sqrt{1+y'^2}}
+ \mu z'(x)\frac{1}{(1+y'^2)^{3/2}}
\right] &=& \di{\phi}{y}\sqrt{1+y'^2}\\
\frac{d}{dx}\left[
z\tanh x + z'\frac{\cosh x+\lambda}{\cosh^3 x}
+ \phi(x,y)\tanh x
+ \frac{\mu z'(x)}{\cosh^3 x}
\right] &=& \di{\phi}{y}\cosh x\nonumber\\
\frac{d}{dx}\left[
z\tanh x + z'(\lambda+\mu)\sech^3x
+z'\sech^2x
+ \phi(x,y)\tanh x
\right] &=& \di{\phi}{y}\cosh x\nonumber
\end{eqnarray}

\begin{equation}
\frac{d}{dx}\left[
z\tanh x + z'(\lambda+\mu)\sech^3(x)
+z'\sech^2x
+ \phi(x,\cosh(x))\tanh(x)
\right] = \di{\phi}{y}\cosh(x)
\end{equation}

Now just write $\lambda$ for $\lambda+\mu$:

\begin{eqnarray}
z'\tanh x + z\sech^2x\nonumber
+
z''\lambda\sech^3x
-z'\cdot 3\lambda\sech^3x\tanh x\\
+
z''\sech^2x
-2z'\cdot\sech^2x\tanh x
+
\phi_x(x,\cosh x) + \sinh x\,\phi_y(x,\cosh x) &=&
\cosh x\,\phi_y(x,\cosh x)
\end{eqnarray}

Collecting terms:

\begin{equation*}
z''\cdot(\lambda\sech^3x+\sech^2x) +
z'\cdot(\tanh x - 3\lambda\sech^3x\tanh x - 2\sech^2 x\tanh x) +
z\cdot(\sech^2x)
=
\phi_x 
+\phi_y(\cosh x-\sinh x)
\nonumber
\end{equation*}

Where we understand that $\phi=\phi(x,y)=\phi(x,\cosh x)$,
$\phi_x=\partial\phi(x,y)/\partial x$ and
$\phi_y=\partial\phi(x,y)/\partial y$, all evaluated at $(x,\cosh x)$.


\begin{equation*}
z''\cdot(\lambda\sech^3x+\sech^2x) +
z'\cdot\tanh x(1 - 3\lambda\sech^3x- 2\sech^2 x) +
z\cdot(\sech^2x)
=
\phi_x 
+\phi_y(\cosh x-\sinh x)
\nonumber
\end{equation*}


Now set $\lambda=0$:

\begin{equation*}
z''\cdot(\sech^2x) +
z'\cdot\tanh x(1 - 2\sech^2 x) +
z\cdot(\sech^2x)
=
\phi_x 
+\phi_y(\cosh x-\sinh x)
\nonumber
\end{equation*}

# Numerical work


```{r odesolver}
library("deSolve") 
sech <- function(x){1/cosh(x)}
Lorenz <- function(t, state, parameters) {
    with(as.list(c(state, parameters)), {
        RHS <- 0.1  # could be dphi/dx + dphi/dy*(cosh(x)-sinh(x))
        dZ <- Zdash
        ss <- sech(t)^2
        dZdash <- (-Z*ss - Zdash*(tanh(t)*(1 - 3*lambda/cosh(t)^3 - 2*ss)) + RHS ) / ( ss + lambda/cosh(t)^3)
        list(c(dZ, dZdash))
    })
}

parameters <- c(lambda = 0.1)
state      <- c(Z = 0, Zdash = -1)  # initial conditions
times      <- seq(-2, 2, by = 0.01)

out <- ode(y = state, times = times, func = Lorenz, parms = parameters)
```

```{r showode}
head(out)
x <- out[,1]
y <- out[,2]
plot(x,y)
grid()
```


```{r defodefunc}
f <- function(lambdamu){
    times <- seq(from = -2, to = 2, by = 1)
    parameters <- c(lambda = lambdamu)
    state      <- c(Z = 0, Zdash = -1)  # initial conditions
    out <- ode(y = state, times = times, func = Lorenz, parms = parameters)
    return(out)
}

f(1)
```
    
