---
title: "Perturbing the Snell law with small changes to refractive index: a first-order analysis"
author: "Robin Hankin"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## The shortest distance between two points is a straight line

\newcommand{\di}[2]{\frac{\partial{#1}}{\partial{#2}}}
\newcommand{\cx}{\cosh{\left(\frac{x-x_0}{a}\right)}}
\newcommand{\sx}{\sinh{\left(\frac{x-x_0}{a}\right)}}
\newcommand{\tx}{\tanh{\left(\frac{x-x_0}{a}\right)}}
\newcommand{\sech}{\operatorname{sech}}


# First, the full equation

\begin{equation}
\frac{d}{dx}\di{F}{y'} -\di{F}{y}=0\qquad F=f(x,y)\sqrt{1+y'^2}
\end{equation}

\begin{equation}
\frac{d}{dx}\left[\frac{y'f(x,y)}{\sqrt{1+y'^2}}\right] = 
\sqrt{1+y'^2}\,f_y(x,y)
\end{equation}


Noting that $\frac{d}{dx}f(x,y)=f_x(x,y) + y'f_y(x,y)$ we have

\begin{equation}
\frac{y''\,f(x,y)}{(1+y'^2)^{3/2}}+
\frac{y'(f_x(x,y)+y'f_y(x,y))}{\sqrt{1+y'^2}} =
\sqrt{1+y'^2}\,f_y(x,y)
\end{equation}

where we understand that $f_y=f_y(x,y)$ etc.

\begin{equation}
y'' + y'(1+y'^2)\frac{f_x(x,y)+y'f_y(x,y)}{f(x,y)} = 
\sqrt{1+y'^2}\,f_y(x,y)
\end{equation}

\begin{equation}
y'' = \frac{f_y-y'f_x}{f}(1+y'^2)
\end{equation}

This equation seems to be not at all easy to solve.  Starting with
$f=f(x,y)=a+bx+cy$ ($a,b,c$ constant) we would have

\begin{equation}
y'' = \frac{c-y'b}{a+bx+cy}(1+y'^2)
\end{equation}

which seems to be extraordinarily difficult to solve.  I guess there
is a trivial solution if $y'=c/b$ [viz $y''=0$]



# Numerical work


```{r odesolver}
library("deSolve") 
Lorenz <- function(t, state, parameters) {
    with(as.list(c(state, parameters)), {
        dZ <- Zdash
        dZdash <- (C - Zdash*B)/(A + B*t + C*Z) * (1+ Zdash^2)

        return(list(c(dZ, dZdash)))
    })
}

parameters <- c(A=1, B=0.1, C=0.1)
state      <- c(Z = 0, Zdash = -1)  # initial conditions
times      <- seq(-2, 54, by = 0.1)

out <- ode(y = state, times = times, func = Lorenz, parms = parameters)
```

```{r showode}
plot(out[,1:2], type='l', asp=1)
for(i in -30:10){abline(i,-1,lwd=0.1)}

```


# Further work:  full E-L but with a small change in refractive index


We start with the Euler-Lagrange equation

\begin{equation}
\frac{d}{dx}\di{F}{y'} -\di{F}{y}=0
\end{equation}


For a uniform refractive index we would have $F=\sqrt{1+y'^2}$ but here we have 

\begin{equation}
F=\sqrt{1+y'^2}(1+\epsilon f(x))
\end{equation}


Where $f(x)$ is some function of $x$.  We will start the path at point
$(-1,-1)$ and end it at $(1,1)$.

First we have $f(x)=0$.  Then E-L gives us

\begin{equation}
\frac{d}{dx}\di{F}{y'}=0
\end{equation}


Thus $\di{F}{y'}=K$ for some constant $K$, or
$\frac{y'}{\sqrt{1+y'^2}}=K$.  For later convenience we will *not*
infer immediately that $y'$ is constant but solve

\begin{equation}
\frac{y'}{\sqrt{1+y'^2}}=K\longrightarrow y'=\frac{K}{\sqrt{1-K^2}}
\end{equation}

This is quite easy to solve.  We have $y'$ is constant and with the
boundary conditions we have $y'=1$ so $K=\sqrt{2}/2$.

Now we perturb the situation with a slight refractive index change:

\begin{equation}
F=\sqrt{1+y'^2}(1+\epsilon H(x))
\end{equation}

Where $\epsilon$ is small and $H(x)$ is the Heaviside step function.

Then we have 

\begin{equation}
\frac{d}{dx}\di{\sqrt{1+y'^2}(1+\epsilon H(x))}{y'}=0
\longrightarrow
\frac{d}{dx}\frac{(1+\epsilon H(x))y'}{\sqrt{1+y'^2}}=0
\longrightarrow
\frac{(1+\epsilon H(x))y'}{\sqrt{1+y'^2}}=K
\end{equation}


So now

\begin{equation}
y'=\frac{K}{\sqrt{(1+\epsilon H(x))^2-K^2}}
\end{equation}

So 

\begin{equation}
y'=\left\{\begin{array}{lr}
\frac{K}{\sqrt{1-K^2}}, & x<0\\
\frac{K}{\sqrt{(1+\epsilon)^2-K^2}}, & x\geqslant 0
\end{array}
\right.
\end{equation}

So, given $y(-1)=-1$ the constraint that the beam passes through $(-1,-1)$ and $(1,1)$ would be 

\begin{equation}
\frac{K}{\sqrt{1-K^2}} + 
\frac{K}{\sqrt{(1+\epsilon)^2 - K^2}} =2
\end{equation}

This is not easy to solve, even if we discard terms of order $\epsilon^2$.
I get

\begin{equation}
\frac{K}{\sqrt{(1+\epsilon)^2 - K^2}} = 
\frac{K}{\sqrt{1-K^2}}\left(   1-\frac{\epsilon}{1-K^2}  \right) + \mathcal{O}(\epsilon^2)
\end{equation}


To first order:

\begin{equation}
\frac{K}{\sqrt{1-K^2}}\left(1 - \frac{\epsilon/2}{1-K^2}  \right)=1
\end{equation}


after algebra

\begin{equation}
2K^4-(3-\epsilon)K^2 +1=0
\end{equation}


from which we recover $K=\sqrt{2}/2$ at $\epsilon=0$, and, to first
order we have $K=\frac{\sqrt{2}}{2}(1+\epsilon/2)$ [the other root is
$1-\epsilon/2$, not sure what that means].  So for $x<0$ we have

\begin{equation}
y'=\frac{K}{\sqrt{1-K^2}}= 1+\epsilon
\end{equation}


and for $x>0$ we have (again to first order)

\begin{equation}
y'=\frac{K}{\sqrt{(1+\epsilon)^2-K^2}}=1-\epsilon
\end{equation}


# Perturbed Euler-Lagrange

OK but look how much easier it is using the perturbed EL:

\begin{equation}
\frac{d}{dx}\left[\di{}{y'}\left(
z\di{F}{y} + z'\di{F}{y'} + G\right)\right] =
\di{}{y}\left(z\di{F}{y} + z'\di{F}{y'} + G\right)
\end{equation}
 
Now we have $F=\sqrt{1+y'^2}$ and $G=\phi\sqrt{1+y'^2}$ where
$\phi=\phi(x,y)$ is the perturbation in refractive index.

We observe that

\begin{equation}
z\di{F}{y} + z'\di{F}{y'} + G = \frac{z'y'}{\sqrt{1+y'^2}} + \phi\sqrt{1+y'^2}
\end{equation}

So we have 


\begin{equation}
\frac{d}{dx}\left[
\frac{z'}{(1+y'^2)^{3/2}} + \frac{\phi y'}{\sqrt{1+y'^2}}
\right] =
\phi_y\sqrt{1+y'^2}
\end{equation}

At this point we remind ourselves that $y'$ is a constant (and we
write $s$ for slope); and also $\frac{d\phi}{dx}=\phi_x + s\phi_y$:



\begin{equation}
\frac{z''}{(1+s^2)^{3/2}}+\frac{y'(\phi_x + s\phi_y)}{\sqrt{1+s^2}}
=
\phi_y\sqrt{1+s^2}
\end{equation}

Simplifying:

\begin{equation}
z'' = (1+s^2)(\phi_y - s\phi_x)
\end{equation}


## Example

Suppose $\phi=ax+by$.  Then we have $\phi_x=a$ and $\phi_y=b$.

\begin{equation}
z'' = (1+s^2)(b-sa)
\end{equation}

This has solution $z=A + Bx + \frac{1}{2}(1+s^2)(b-sa)x^2$ where $A$
and $B$ are constants of integration.

### Full E-L equations for 

OK now we use the Full EL equations, starting from Beltrami:

\begin{equation}
F -y'\di{F}{y'}=C,\qquad F=f(y)\sqrt{1+y'^2}
\end{equation}

\begin{equation}
f(y)\sqrt{1+y'^2} - \frac{f(y)y'^2}{\sqrt{1+y'^2}} = C
\end{equation}

\begin{equation}
f(y)(1+y'^2) - f(y)y'^2 = C\sqrt{1+y'^2}
\end{equation}

\begin{equation}
f(y) = C\sqrt{1+y'^2}
\end{equation}

\begin{equation}
k^2f(y)^2 = 1+y'^2
\end{equation}

\begin{equation}
y'=\sqrt{k^2f(y)^2-1}
\end{equation}

\begin{equation}
\int dx=\int\frac{dy}{\sqrt{k^2f(y)^2-1}}
\end{equation}

This integral is difficult for any non-trivial $f(\cdot)$.  For
example, if $f(y)=1+ay^2$, mathematica gives

\begin{equation}
-\frac{i \sqrt{\frac{a k y^2}{k-1}+1} \sqrt{\frac{a k
   y^2}{k+1}+1} F\left(i \sinh ^{-1}\left(\sqrt{\frac{a k}{k+1}}
   y\right)|\frac{k+1}{k-1}\right)}{\sqrt{\frac{a k}{k+1}}
   \sqrt{\left(a k y^2+k\right)^2-1}}
\end{equation}

where $F=F(\phi\vert m)$ is the elliptic integral of the first kind;
this is a bit of a nightmare to work with (note that the expression is
complex).  Note that we have $x$ in terms of $y$.

Anyway

# Perturbed Euler-Lagrange but with $1+\epsilon H(x)$

\begin{equation}
\frac{d}{dx}\left[\di{}{y'}\left(
z\di{F}{y} + z'\di{F}{y'} + G\right)\right] =
\di{G}{y}
\end{equation}

Here we have $F=\sqrt{1+y'^2}$ and $G=H(x)\sqrt{1+y'^2}$.


\begin{equation}
\frac{d}{dx}\left[\di{}{y'}\left(
\frac{z'y'}{\sqrt{1+y'^2}}
+ H(x)\sqrt{1+y'^2}
\right)\right] =
0
\end{equation}


\begin{equation}
\di{}{y'}\left(
\frac{z'y'}{\sqrt{1+y'^2}}
+ H(x)\sqrt{1+y'^2}
\right)
= K
\end{equation}

\begin{equation}
\frac{z'}{(1+y'^2)^{3/2}}
+ \frac{y'H(x)}{\sqrt{1+y'^2}}
= K
\end{equation}



\begin{equation}\label{tobegeneralized}
\frac{z'}{(1+y'^2)^{3/2}}
+ \frac{y'H(x)}{\sqrt{1+y'^2}}
= K
\end{equation}

Now the clever bit, we know $y'=1$:

\begin{equation}
z'\frac{\sqrt{2}}{4}
+ \frac{\sqrt{2}}{2}H(x)
= K\longrightarrow z'=2\sqrt{2}K-2H(x)
\end{equation}

So the constraint for $K$ becomes $2\sqrt{2}K + 2\sqrt{2}K -2=0$, or
$K=\sqrt{2}/4$ (NB there is no sense in which we wish to "recover"
$K=\sqrt{2}/2$.  This $K$ is the constant of integration for the
perturbation, not the zeroth order straight line problem).

\begin{equation}
z'=1-2H(x)=\left\{\begin{array}{lr}1&x<0\\ -1&x\geqslant 0\end{array}\right.
\end{equation}


# Now something a little harder

A slight generalization:  We 
constrain the path to travel from $(x_1,y_1)$ to $(x_2,y_2)$, and have 
a more general variation of refractive index, say $1+\epsilon f(x)$.  


So now $F=\sqrt{1+y'^2}$, $G=\sqrt{1+y'^2}f(x)$

\begin{equation}
\frac{d}{dx}\left[\di{}{y'}\left(
z\di{F}{y} + z'\di{F}{y'} + G\right)\right] =
\di{G}{y}
\end{equation}

Right hand side is zero.  We have 

\begin{equation}
\frac{d}{dx}\left[z'\frac{\partial^2F}{\partial y'^2}
+ \di{G}{y'}\right] = 0
\end{equation}

\begin{equation}
z'\frac{\partial^2F}{\partial y'^2}
+ \di{G}{y'}=K
\end{equation}


\begin{equation}
\frac{z'}{(1+y'^2)^{3/2}}  + \frac{y'f(x)}{\sqrt{1+y'^2}} = K
\end{equation}


Now we know that $y'=(y_2-y_1)/(x_2-x_1)$.  We will write this as $s$
for slope.

\begin{equation}
\frac{z'}{(1+s^2)^{3/2}}  + \frac{s\,f(x)}{\sqrt{1+s^2}} = K
\end{equation}


\begin{equation}
z' = K(1+s^2)^{3/2} - s(1+s^2)f(x)
\end{equation}

We know that $z(x_1)=0$ [which gives us the constant of integration]
and $z(x_2)=0$ [which gives us $K$].  Observing that $s=(y_2-y_1)/(x_2-x_1)$ we have

\begin{equation}
z(x_2) = 0 = 0 + (x_2-x_1)K(1+s^2)^{3/2} -
s(1+s^2)\int_{x_1}^{x_2}f(x)dx
\end{equation}

\begin{equation}
K = \frac{s\overline{\mathcal{I}}}{\sqrt{1+s^2}}
\end{equation}

(where
$\overline{\mathcal{I}}=\frac{\int_{x_1}^{x_2}f(x)\,dx}{x_2-x_1}$ is
the mean integral of $f(x)$ from $x_1$ to $x_2$).

Substituting:


\begin{equation}
z' = \overline{\mathcal{I}}s(1+s^2) - s(1+s^2)f(x) 
\end{equation}

Simplifying:

\begin{equation}\label{final}
z' = s(1+s^2)(\overline{\mathcal{I}}-f(x))
\end{equation}

\section{Now a function of y}

Will use $F = f(y)\sqrt{1+y'^2}$.


\begin{equation}
\frac{d}{dx}\di{F}{y'}=\di{F}{y}
\end{equation}

\begin{equation}\label{covstraight}
\frac{d}{dx}\left(\frac{y'f(y)}{\sqrt{1+y'^2}}\right)
= f'(y)\sqrt{1+y'^2}
\end{equation}

Now we need that 

\begin{eqnarray*}
\frac{d f(y)}{dx} &=& \frac{d f(y(x))}{dx}\\
&=& \frac{f(y(x+\delta x))-f(y(x))}{\delta x}\\
&=& \frac{f(y(x)+y'(x)\delta x))-f(y(x))}{\delta x}\\
&=& \frac{f(y(x)) + y'(x)\delta xf'(y(x))-f(y(x))}{\delta x}\\
&=& y'(x)f'(y(x))
\end{eqnarray*}


Thus 
\begin{equation}
\frac{d}{dx}\left(\frac{y'f(y)}{\sqrt{1+y'^2}}\right) =
\frac{y''f(y)}{(1+y'^2)^{3/2}} + \frac{y'^2f'(y)}{\sqrt{1+y'^2}}
\end{equation}


Equation~\ref{covstraight} becomes

\begin{equation}
\frac{y''f(y)}{(1+y'^2)^{3/2}} + \frac{y'^2f'(y)}{\sqrt{1+y'^2}} =
f'(y)\sqrt{1+y'^2}
\end{equation}

\begin{equation}
y''f(y) + y'^2(1+y'^2)f'(y) = f'(y)(1+y'^2)^2
\end{equation}

\begin{equation}
y'' = \frac{f'(y)}{f(y)}(1+y'^2) = d\log(f(y))/dy(1+y'^2)
\end{equation}


We an take, as an example, $d\log(f(y))/dy$ to be a constant $K$:

\begin{equation}
y'' = K(1+y'^2)
\end{equation}



Use the linearization:

\begin{equation}
\frac{d}{dx}\left[\di{}{y'}\left(
z\di{F}{y} + z'\di{F}{y'} + G\right)\right] =
\di{G}{y}
\end{equation}


Here $F=\sqrt{1+y'^2}$ and $G=f(y)\sqrt{1+y'^2}$ so we have 


\begin{equation}
\frac{d}{dx}\left[\di{}{y'}\left(
z'\di{\sqrt{1+y'^2}}{y'} + f(y)\sqrt{1+y'^2}\right)\right] =
\di{f(y)\sqrt{1+y'^2}}{y}
\end{equation}


\begin{equation}
\frac{d}{dx}
\left[
\frac{y'z'}{\sqrt{1+y'^2}} + \frac{y'f(y)}{\sqrt{1+y'^2}}
\right]
=
f'(y)\sqrt{1+y'^2}
\end{equation}

\begin{equation}
\frac{d}{dx}
\left[
\frac{y'(z'+f'(y))}{\sqrt{1+y'^2}}
\right]
=
f'(y)\sqrt{1+y'^2}
\end{equation}

writing $s=y'$ for the slope of the unperturbed line

\begin{equation}
\frac{d}{dx}
\left[
(z'+f'(y))\frac{s}{\sqrt{1+s^2}}
\right]
=
f'(y)\sqrt{1+s^2}
\end{equation}

\begin{equation}
z''+sf''(y)
=
f'(y)\frac{1+s^2}{s}
\end{equation}



\section{Use of the Beltrami equation}

Beltrami, viz:

\begin{equation}
F-u'\di{F}{y'}=C
\end{equation}

is arguably a bit easier to apply.  We apply the same perturbation as
before, specifically $F\longrightarrow F+\epsilon G$.  Taking the
first order terms and subtracting out Beltrami we get

\begin{equation}\label{beltrami_epsilon}
G-y'\di{G}{y'}-z'\di{F}{y'}=C
\end{equation}

for some (other) constant $C$.  Now perturb it with $G=f(y)$ and
substitute into equation \ref{beltrami_epsilon}:

\begin{equation}
f(y) - \frac{z'y'}{\sqrt{1+y'^2}}=C
\end{equation}

\begin{equation}
f(y) - \frac{z's}{\sqrt{1+s^2}}=C
\end{equation}



